---
title: "Introduction to R, part 2a"
author: "Steve Simon"
date: "Saturday, August 13, 2016"
output: html_document
---

This program is part of a series of R programs for the Introduction to R class. This program shows you how to import a data set on mortality aboard the Titanic.

```{r r-version-and-current-date}
R.version.string
Sys.Date()
```

Start with a clean slate.

```{r clean-slate}
rm(list=ls())
```

You should find the data file on mortality among passengers of the Titanic at

--> http://www.statsci.org/data/general/titanic.txt

A brief description of this file is available at

--> http://www.statsci.org/data/general/titanic.html

You need to read in the file using the read.table function.

```{r read, error=TRUE}
fn <- "http://www.statsci.org/data/general/titanic.txt"
ti <- read.table(file=fn)
```

There is an issue on the 38th line of the file. You need to download the file and fix it and then read it locally.

First, read the original file to confirm the error.

```{r read1a, error=TRUE}
fn <- "titanic_v00.txt"
ti <- read.table(file=fn)
```

We removed some double quote marks from line 38: Brown, Mrs James Joseph (Margaret Molly Tobin).

```{r read2, error=TRUE}
fn <- "titanic_v01.txt"
ti <- read.table(file=fn)
```

Oops! There is one more problem. Quote marks needed to be added to line 458: Jacobsohn Mr Sidney Samuel. Fix this, save under a different name and then read it in.

```{r read3, error=TRUE}
fn <- "titanic_v02.txt"
ti <- read.table(file=fn)
head(ti)
tail(ti)
```

Darn it all, we almost have it. But the variable names are listed as the first row of the data set. Take a look at the help file for the read.table function to see how to fix this.

```{r read4, error=TRUE}
ti <- read.table(file=fn, header=TRUE)
head(ti)
tail(ti)
```

Just for the record, you can read the original file without any difficulty by adding an option under the "sep" argument. The default is to separate by white space (blanks or tabs, mostly), but when some of the quoted names are not quoted properly, the spaces in the names throw things off. If you separate just by tabs and not by tabs and/or spaces, everything works perfectly.

I only found this out, by the way, through trial and error.

```{r read6, error=TRUE}
fn <- "http://www.statsci.org/data/general/titanic.txt"
ti <- read.table(file=fn, sep="\t", header=TRUE)
```


Notice that Miss Tamini Zabour has an "NA" for age.  We'll take a careful look at this later.

## On your own (you do not need to turn this in)

Who is the oldest passenger on the Titanic. Did he/she survive?

## Categorical versus continuous variables

A categorical variable is defined (loosely) as a variable that has a small number of possible values. Each value is usually associated with a particular category or label. In contrast, a continuous variable is defined as a variable that has a large number of possible values, potentially any value in a particular interval.

Yesterday, almost all of the variables that you used were continuous. Today, almost all of the variables that you will use will be categorical.

The distinction between continuous and categorical variables is important in deciding what types of descriptive and inferential statistics you should use. But, there is often gray and fuzzy line between categorical and continuous variables. Don't worry too much about this today. If you're not sure whether a variable is categorical or continuous, try some simple descriptive statistics and graphs appropriate for categorical data and then try some simple descriptive statistics and graphs for continuous data. You will usually figure out quickly whehther treating your variable as categorical or continuous makes the most sense.

There are other types of variables also, such as count variables, that have their own special features.

## Save all my work

```{r save-image}
save.image("part2.RData")
```

Please take the time to answer the quiz found at

http://www.pmean.com/15/images/quiz2a.html