---
title: "Introduction to R, part 2"
author: "Steve Simon"
date: "Saturday, August 13, 2016"
output: html_document
---

This program illustrates how to import and manipulate simple data sets in R and how to produce some simple descriptive statistics. Included in this program are some exercises for you to do on your own.

# Par 2d. Bar plots

Let's erase any stray data and load the data from part 2c.

```{r read-part-2c-data}
rm(list=ls())
load(file="part2c.RData")
```

I'm not a big fan of bar plots, but they sometimes have their uses. You can get a barplot for the frequency count

```{r barplot}
barplot(table(ti$PClass))
```

or for the percentage (note the use of the ylab argument here).

```{r barplot2}
barplot(100*prop.table(table(ti$PClass)),ylab="Percent")
```

You can make things even nicer by pasting a percent sign on the vertical axis tick marks.

```{r barplot3}
barplot(100*prop.table(table(ti$PClass)),axes=FALSE)
axis(side=2,at=10*(0:5),labels=paste(10*(0:5),"%",sep=""))
```

## On your own

Add label "Passenger class" to the x-axis of this bar plot.

## Save all my work

```{r save-image}
save.image("part2d.RData")
```