---
title: "Introduction to R, part 2e"
author: "Steve Simon"
date: "Saturday, August 13, 2016"
output: html_document
---

This program is part of a series of R programs for the Introduction to R class. This program shows you how to create two dimensional crosstabulations.

```{r r-version-and-current-date}
R.version.string
Sys.Date()
```
Let's erase any stray data and load the data from part 2d.

```{r read-part-2d-data}
ls()
rm(list=ls())
load(file="part2.RData")
ls()
```

When you are looking at the relationship between two categorical variables, you need to start with a crosstabulation. You use the table function in R with two arguments to create a crosstabulation.

```{r crosstabs}
table(ti$PClass,ti$surv.factor)
```

The first variable is the rows and the second variable is the columns.

As a general rule of thumb, if one variable tends to have a lot of  categories, I tend to put that in the rows. This allows you to scan the many categories with an up/down scroll rather than a left/right scroll. There are other rules of thumb for which variable to choose for your rows and which to choose for your columns in a crosstabulation.

Again, notice that R is rather minimalist in its approach. You do not see any row or column totals. You can get these with the addmargins function.

```{r add.margins}
ct <- table(ti$PClass,ti$surv.factor)
addmargins(ct,1)
addmargins(ct,2)
addmargins(ct)
```

## Table percentages.

Percentages in a crosstabulation are actually quite tricky. You can compute a percentage across the entire table (which would be called a cell percent in SPSS and SAS). It helps to round these results.

```{r cell}
prop.table(ct)
round(100*prop.table(ct))
```

These percentages add to 100% across all of the table. In contrast, row percentages add up to 100% within each row.

```{r row}
round(100*prop.table(ct,margin=1))
```

This is a rather cryptic command and it shows one area for R where getting simple statistics is a lot harder than in SAS or SPSS.

Column percentages add up to 100% within each column.

```{r column}
round(100*prop.table(ct,margin=2))
```

It is not always clear whether you should use row, column, or cell percents. You can find some general advice at

--> http://www.pmean.com/02/percentage.html

but to be honest, the best approach is to try several approaches and see what works best. I have found that I usually like row percentages in a crosstabulation with your outcome variables as the columns.

## On your own

Compute a table showing the relationship between sex and survival. The Titanic sunk during a time when people really did believe in the phrase "women and children first". Find a table of percentages that shows the degree to which women were more likely to survive than men.

## Save all my work

```{r save-image}
save.image("part2.RData")
```

Please take the time to answer the quiz found at

http://www.pmean.com/15/images/quiz2e.html