---
title: "Introduction to R, part 2d"
author: "Steve Simon"
date: "Saturday, August 13, 2016"
output: html_document
---

This program is part of a series of R programs for the Introduction to R class. This program shows you how to draw a bar chart.

```{r r-version-and-current-date}
R.version.string
Sys.Date()
```

Let's erase any stray data and load the data from part 2c.

```{r read-part-2c-data}
ls()
rm(list=ls())
load(file="part2.RData")
ls()
```

I'm not a big fan of bar plots, but they sometimes have their uses. You can get a barplot for the frequency count

```{r barplot}
barplot(table(ti$PClass))
```

or for the percentage (note the use of the ylab argument here).

```{r barplot2}
barplot(100*prop.table(table(ti$PClass)),ylab="Percent")
```

You can make things even nicer by pasting a percent sign on the vertical axis tick marks.

```{r barplot3}
barplot(100*prop.table(table(ti$PClass)),axes=FALSE)
axis(side=2,at=10*(0:5),labels=paste(10*(0:5),"%",sep=""))
```

## On your own (you do not need to turn this in)

Add label "Passenger class" to the x-axis of this bar plot.

## Save all my work

```{r save-image}
save.image("part2.RData")
```

Please take the time to answer the quiz found at

http://www.pmean.com/15/images/quiz2d.html