---
title: "Introduction to R, part 2b"
author: "Steve Simon"
date: "Saturday, August 13, 2016"
output: html_document
---

This program is part of a series of R programs for the Introduction to R class. This program shows you how to get frequency counts for a categorical variable.

```{r r-version-and-current-date}
R.version.string
Sys.Date()
```

Erase any stray data and load the data from part 2a.

```{r read-part2-data}
ls()
rm(list=ls())
load(file="part2.RData")
ls()
```

For categorical variables, you should first get frequency counts.

```{r table-simple}
table(ti$PClass)
```

Unlike most other statistical packages, R tends to have a minimalist approach to statistics. So the table function produces only counts. If you want percentages in addition to counts, there are several approaches.

The prop.table function takes a frequency table and converts it to a proportion.

```{r table-proportion}
tb <- table(ti$PClass)
prop.table(tb)
```

Multiply by 100 and round to get percentages.

```{r table-pct}
pct <- round(100*prop.table(tb))
pct
```

You can get even fancier. The paste function concatenates several string variables and if something is not a string, R will convert it to a string before concatenating.

```{r table-pct-sign}
pct.sign <- "%"
paste(pct,pct.sign,sep="")
```

Somehow, the names of the passenger class got lost. So let's add them back in.

```{r table-names}
paste(names(pct),pct,pct.sign,sep="")
```

You can add spacing and a bit of punctuation to make the output look more readable.

```{r table-punctuation}
colon <- ": "
paste(names(tb),colon,pct,pct.sign,sep="")
```

Finally, you can combine the counts and the total with the percents to make things look really nice.

```{r fractions}
n  <- sum(tb)
slash <- "/"
comma <- ", "
paste(names(tb),colon,tb,slash,n,comma,pct,pct.sign,sep="")
```

This shows how I like to program in R. First, get something simple. Then slowly add layers to it until you get a nice polished product.

In contrast, most other statistical packages try to produce polished results right from the start. This has some advantages, but the approach used by R, where many of the functions do something simple and basic, allows you to polish the results the way you want them.

# Tables that include counts of missing values.

The default option in the table function is to not list missing values. During the intitial data screening, you should always look for missing values. This is done with the useNA argument.

```{r table-usena}
table(ti$PClass,useNA="always")
```

You can add a total count to the frequency table with the addmargins function.

```{r table-addmargins}
addmargins(table(ti$PClass,useNA="always"))
```

## On your own

Run frequency counts for Sex and Survived. Be sure to look for missing values for these two variables.

## Save all my work

```{r save-image}
save.image("part2b.RData")
```

Please take the time to answer the quiz found at

http://www.pmean.com/15/images/quiz2b.html